---
title: "Are Bangalore Summers Getting Hotter?"
subtitle: "What the data actually says vs what Twitter says"
output: html_notebook
---

```{r setup, message=FALSE, warning=FALSE}
require(tidyverse)
require(lubridate)
require(ggthemes)
require(ggrepel)
require(patchwork)

load('/Users/Karthik/Documents/work/data work/bangalore/weather/data/bangaloreTemperature.RData')

blrTemp %>%
  filter(!is.na(Temp)) %>%
  mutate(
    Temp = as.numeric(Temp),
    Hour = hour(DT),
    Year = year(DT),
    Month = month(DT),
    Day = day(DT),
    Date = as.Date(DT)
  ) %>%
  filter(Year >= 1981, Year <= 2025) ->
  blrTemp

# Reusable theme
theme_blr <- function() {
  ggthemes::theme_tufte() +
    theme(
      panel.grid = element_blank(),
      panel.background = element_rect(fill = '#e5e1d8', linewidth = 0),
      plot.background = element_rect(fill = '#e5e1d8', linewidth = 0),
      axis.text = element_text(face = 'bold', color = '#3C3C3C'),
      axis.line.y = element_line(colour = 'black', linewidth = 0.2),
      axis.ticks.x = element_blank(),
      legend.position = 'none',
      plot.title = element_text(face = 'bold', color = '#3C3C3C'),
      plot.subtitle = element_text(size = 9, color = '#5f3946'),
      plot.caption = element_text(color = '#3C3C3C'),
      plot.margin = margin(15, 15, 15, 15)
    )
}

# Summer = March, April, May
summer <- blrTemp %>% filter(Month %in% 3:5)

# Daily summaries
summer_daily <- summer %>%
  summarise(High = max(Temp), Low = min(Temp), Mean = mean(Temp), .by = c(Date, Year))
```


## The headline number: are summer highs going up?

The most common claim on Twitter: "it used to never cross 35, now it does all the time." Let's check.

```{r summer-highs-trend, fig.width=10, fig.height=5}
summer_daily %>%
  summarise(
    MeanHigh = mean(High),
    MaxHigh = max(High),
    .by = Year
  ) ->
  yearly_highs

lm_mean <- lm(MeanHigh ~ Year, data = yearly_highs)
slope_mean <- round(coef(lm_mean)[2] * 10, 2)

# Label outliers
yearly_highs %>%
  mutate(
    Label = case_when(
      MeanHigh == max(MeanHigh) ~ paste0(Year, ": ", round(MeanHigh, 1), "\u00B0C"),
      MeanHigh == min(MeanHigh) ~ paste0(Year, ": ", round(MeanHigh, 1), "\u00B0C"),
      MaxHigh == max(MaxHigh) ~ paste0(Year, " peak: ", round(MaxHigh, 1), "\u00B0C"),
      TRUE ~ ""
    )
  ) ->
  yearly_highs

yearly_highs %>%
  ggplot(aes(x = Year, y = MeanHigh)) +
  geom_point(color = "#5f3946", size = 2) +
  geom_smooth(method = 'lm', se = FALSE, color = "#CD3333", linewidth = 0.7, linetype = 'dashed') +
  geom_text_repel(aes(label = Label), size = 3, fontface = 'bold', color = '#3C3C3C',
                  nudge_y = 0.3, min.segment.length = 0) +
  annotate("text", x = 1995, y = 34.5,
           label = paste0("Trend: +", slope_mean, "\u00B0C per decade"),
           size = 3.5, fontface = 'bold', color = '#CD3333') +
  scale_y_continuous(breaks = seq(31, 35, 0.5)) +
  theme_blr() +
  labs(
    title = "Average Summer Daily High: Barely Moving",
    subtitle = "Mean of daily maximum temperature, March-May each year. The trend is modest.",
    x = "", y = "Mean daily high (\u00B0C)",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```

The average summer afternoon is warming at about +0.1 C per decade. That's real but small - you wouldn't feel the difference between 1985 and 2025 on a typical day.


## But the extremes are a different story

The number of days crossing 35 C tells a sharper story.

```{r extreme-days, fig.width=10, fig.height=5}
summer_daily %>%
  summarise(
    Above35 = sum(High >= 35),
    Above36 = sum(High >= 36),
    .by = Year
  ) ->
  extreme_counts

extreme_counts %>%
  ggplot(aes(x = Year, y = Above35)) +
  geom_col(fill = "#5f3946", width = 0.7) +
  geom_text(
    data = extreme_counts %>% filter(Above35 >= 15),
    aes(label = Above35), vjust = -0.5, size = 3, fontface = 'bold', color = '#CD3333'
  ) +
  geom_smooth(method = 'loess', se = FALSE, color = "#CD3333", linewidth = 0.7, span = 0.5) +
  annotate("text", x = 2024, y = 38,
           label = "2024:\n35 days",
           size = 3, fontface = 'bold', color = '#CD3333', lineheight = 0.9) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_blr() +
  labs(
    title = "Days Above 35\u00B0C Per Summer: The Spikes Are Getting Bigger",
    subtitle = "Number of days where the daily max exceeded 35\u00B0C, March-May. 2024 was unprecedented.",
    x = "", y = "Days above 35\u00B0C",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```


## The real shift: summer nights

Here's what people might actually be responding to - even if they describe it as "hotter summers." You feel heat not just at 3 PM, but when you can't cool down at night.

```{r summer-nights, fig.width=10, fig.height=5}
summer_daily %>%
  summarise(MeanLow = mean(Low), .by = Year) ->
  yearly_lows

lm_low <- lm(MeanLow ~ Year, data = yearly_lows)
slope_low <- round(coef(lm_low)[2] * 10, 2)

yearly_lows %>%
  mutate(
    Label = case_when(
      MeanLow == max(MeanLow) ~ paste0(Year, ": ", round(MeanLow, 1), "\u00B0C"),
      MeanLow == min(MeanLow) ~ paste0(Year, ": ", round(MeanLow, 1), "\u00B0C"),
      TRUE ~ ""
    )
  ) ->
  yearly_lows

yearly_lows %>%
  ggplot(aes(x = Year, y = MeanLow)) +
  geom_point(color = "#2255AA", size = 2) +
  geom_smooth(method = 'lm', se = FALSE, color = "#CD3333", linewidth = 0.7, linetype = 'dashed') +
  geom_text_repel(aes(label = Label), size = 3, fontface = 'bold', color = '#3C3C3C',
                  min.segment.length = 0) +
  annotate("text", x = 1995, y = 22.5,
           label = paste0("Trend: +", slope_low, "\u00B0C per decade"),
           size = 3.5, fontface = 'bold', color = '#CD3333') +
  theme_blr() +
  labs(
    title = "Summer Nights Are Warming 3x Faster Than Afternoons",
    subtitle = paste0("Mean daily minimum temperature, March-May. Nights warming at +", slope_low, "\u00B0C/decade vs +", slope_mean, "\u00B0C/decade for highs."),
    x = "", y = "Mean daily low (\u00B0C)",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```


## Afternoon vs night: the diverging trends

Let's put both on the same chart to make the asymmetry visible.

```{r afternoon-vs-night, fig.width=10, fig.height=5}
summer %>%
  mutate(
    TimeSlot = case_when(
      Hour %in% 13:16 ~ "Afternoon (1-4 PM)",
      Hour %in% 2:5 ~ "Night (2-5 AM)",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(TimeSlot)) %>%
  summarise(MeanTemp = mean(Temp), .by = c(Year, TimeSlot)) ->
  slot_yearly

# Compute slopes for labels
slopes <- slot_yearly %>%
  group_by(TimeSlot) %>%
  summarise(
    Slope = round(coef(lm(MeanTemp ~ Year))[2] * 10, 2),
    LastTemp = MeanTemp[Year == max(Year)],
    .groups = 'drop'
  )

slot_yearly %>%
  ggplot(aes(x = Year, y = MeanTemp, color = TimeSlot)) +
  geom_point(size = 1.5, alpha = 0.6) +
  geom_smooth(method = 'lm', se = FALSE, linewidth = 0.8, linetype = 'dashed') +
  # Direct labels at end
  geom_text(
    data = slot_yearly %>% filter(Year == 2025),
    aes(label = TimeSlot),
    hjust = -0.05, size = 3, fontface = 'bold', nudge_x = 0.5
  ) +
  scale_color_manual(values = c("Afternoon (1-4 PM)" = "#CD3333", "Night (2-5 AM)" = "#2255AA")) +
  scale_x_continuous(limits = c(1981, 2032)) +
  annotate("text", x = 1990, y = 31,
           label = paste0("Afternoon: +", slopes$Slope[slopes$TimeSlot == "Afternoon (1-4 PM)"], "\u00B0C/decade"),
           color = "#CD3333", fontface = 'bold', size = 3, hjust = 0) +
  annotate("text", x = 1990, y = 22.5,
           label = paste0("Night: +", slopes$Slope[slopes$TimeSlot == "Night (2-5 AM)"], "\u00B0C/decade"),
           color = "#2255AA", fontface = 'bold', size = 3, hjust = 0) +
  theme_blr() +
  labs(
    title = "Two Different Warming Stories at Two Times of Day",
    subtitle = "Summer afternoon temperatures are nearly flat; summer nights are climbing steadily",
    x = "", y = "Mean temperature (\u00B0C)",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```


## Heat streaks: consecutive days above 35 C

One bad day is an anecdote. A streak of bad days is what makes summers feel unbearable.

```{r heat-streaks, fig.width=10, fig.height=5}
summer_daily %>%
  arrange(Year, Date) %>%
  group_by(Year) %>%
  summarise(
    MaxStreak = {
      r <- rle(High >= 35)
      max(c(r$lengths[r$values], 0))
    },
    .groups = 'drop'
  ) ->
  streaks

streaks %>%
  mutate(
    Label = case_when(
      MaxStreak >= 10 ~ paste0(MaxStreak, " days"),
      TRUE ~ ""
    )
  ) %>%
  ggplot(aes(x = Year, y = MaxStreak)) +
  geom_col(fill = "#5f3946", width = 0.7) +
  geom_text(aes(label = Label), vjust = -0.3, size = 3, fontface = 'bold', color = '#CD3333') +
  geom_hline(
    yintercept = mean(streaks$MaxStreak[streaks$Year <= 2000]),
    linetype = 'dashed', color = '#9c9280', linewidth = 0.5
  ) +
  annotate("text", x = 1987, y = mean(streaks$MaxStreak[streaks$Year <= 2000]) + 0.8,
           label = paste0("1981-2000 avg: ", round(mean(streaks$MaxStreak[streaks$Year <= 2000]), 1), " days"),
           size = 2.8, fontface = 'bold', color = '#9c9280') +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_blr() +
  labs(
    title = "Longest Heat Streak Each Summer (Consecutive Days Above 35\u00B0C)",
    subtitle = "2024's 23-day streak was unlike anything in the historical record",
    x = "", y = "Consecutive days",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```


## The daily cycle in summer: then vs now

How does a "typical summer day" in the 1980s compare to one now, hour by hour?

```{r diurnal-shift, fig.width=10, fig.height=5}
summer %>%
  mutate(
    Era = case_when(
      Year <= 1995 ~ "1981-1995",
      Year >= 2011 ~ "2011-2025",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Era)) %>%
  summarise(MeanTemp = mean(Temp), .by = c(Hour, Era)) ->
  diurnal_era

# Compute the gap
diurnal_era %>%
  pivot_wider(names_from = Era, values_from = MeanTemp) %>%
  mutate(Diff = `2011-2025` - `1981-1995`) ->
  diurnal_diff

max_diff <- diurnal_diff %>% slice_max(Diff, n = 1)
min_diff <- diurnal_diff %>% slice_min(Diff, n = 1)

p1 <- diurnal_era %>%
  ggplot(aes(x = Hour, y = MeanTemp, color = Era)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.5) +
  geom_text(
    data = diurnal_era %>% filter(Hour == 14),
    aes(label = Era), hjust = -0.1, size = 3, fontface = 'bold', nudge_y = c(0.3, -0.3)
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3), labels = paste0(seq(0, 23, 3), ":00")) +
  scale_color_manual(values = c("1981-1995" = "#9c9280", "2011-2025" = "#CD3333")) +
  theme_blr() +
  labs(title = "Average Summer Day: 1981-1995 vs 2011-2025", x = "", y = "\u00B0C")

p2 <- diurnal_diff %>%
  ggplot(aes(x = Hour, y = Diff)) +
  geom_col(fill = "#5f3946", width = 0.7) +
  geom_text(aes(label = paste0("+", round(Diff, 1))), vjust = -0.3, size = 2.5, fontface = 'bold', color = '#3C3C3C') +
  geom_text(
    data = max_diff,
    aes(label = paste0("Peak shift:\n+", round(Diff, 1), "\u00B0C at ", Hour, ":00")),
    vjust = -1.5, size = 3, fontface = 'bold', color = '#CD3333', lineheight = 0.9
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3), labels = paste0(seq(0, 23, 3), ":00")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  theme_blr() +
  labs(title = "How Much Warmer at Each Hour?", x = "", y = "Difference (\u00B0C)")

p1 + p2 +
  plot_layout(ncol = 1, heights = c(60, 40)) +
  plot_annotation(
    subtitle = "The gap is largest at night and smallest in the afternoon",
    caption = "Data: ERA5 via Oikolab"
  ) &
  theme(
    plot.subtitle = element_text(size = 9, color = '#5f3946'),
    panel.background = element_rect(fill = "#e5e1d8"),
    plot.background = element_rect(fill = "#e5e1d8", linewidth = 0)
  )
```


## Month-by-month: which summer month is warming fastest?

```{r monthly-trend, fig.width=10, fig.height=7}
summer_daily %>%
  mutate(Month = month(Date)) %>%
  summarise(
    MeanHigh = mean(High),
    MeanLow = mean(Low),
    .by = c(Year, Month)
  ) %>%
  pivot_longer(c(MeanHigh, MeanLow), names_to = "Metric", values_to = "Temp") %>%
  mutate(
    MonthName = month.name[Month],
    MetricLabel = ifelse(Metric == "MeanHigh", "Daily high", "Daily low")
  ) ->
  monthly_trends

# Compute slopes
monthly_trends %>%
  group_by(MonthName, MetricLabel) %>%
  summarise(
    Slope = round(coef(lm(Temp ~ Year))[2] * 10, 2),
    LastY = max(Year),
    LastTemp = Temp[Year == max(Year)],
    .groups = 'drop'
  ) ->
  month_slopes

monthly_trends %>%
  mutate(MonthName = factor(MonthName, levels = c("March", "April", "May"))) %>%
  ggplot(aes(x = Year, y = Temp, color = MetricLabel)) +
  geom_point(size = 0.8, alpha = 0.4) +
  geom_smooth(method = 'lm', se = FALSE, linewidth = 0.8) +
  facet_wrap(~MonthName, nrow = 1) +
  scale_color_manual(values = c("Daily high" = "#CD3333", "Daily low" = "#2255AA")) +
  theme_blr() +
  theme(
    legend.position = 'bottom',
    legend.title = element_blank(),
    strip.text = element_text(face = 'bold', hjust = 0, size = 11)
  ) +
  labs(
    title = "Summer Warming by Month: Highs vs Lows",
    subtitle = "In all three months, nightly lows are climbing faster than daytime highs",
    x = "", y = "Temperature (\u00B0C)",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```

```{r monthly-slopes-table}
month_slopes %>%
  pivot_wider(names_from = MetricLabel, values_from = c(Slope, LastTemp)) %>%
  select(MonthName, `Slope_Daily high`, `Slope_Daily low`) %>%
  rename(Month = MonthName, `High trend (C/decade)` = `Slope_Daily high`, `Low trend (C/decade)` = `Slope_Daily low`) ->
  slope_table

knitr::kable(slope_table, caption = "Warming rate per decade, by month and metric")
```


## The diurnal range is shrinking

If nights warm faster than days, the gap between daily high and daily low should be narrowing. This is a classic urban heat island signature.

```{r diurnal-range, fig.width=10, fig.height=5}
summer_daily %>%
  mutate(Range = High - Low) %>%
  summarise(MeanRange = mean(Range), .by = Year) ->
  yearly_range

lm_range <- lm(MeanRange ~ Year, data = yearly_range)
slope_range <- round(coef(lm_range)[2] * 10, 2)

yearly_range %>%
  mutate(
    Label = case_when(
      MeanRange == max(MeanRange) | MeanRange == min(MeanRange) ~ paste0(Year, ": ", round(MeanRange, 1), "\u00B0C"),
      TRUE ~ ""
    )
  ) %>%
  ggplot(aes(x = Year, y = MeanRange)) +
  geom_point(color = "#5f3946", size = 2) +
  geom_smooth(method = 'lm', se = FALSE, color = "#CD3333", linewidth = 0.7, linetype = 'dashed') +
  geom_text_repel(aes(label = Label), size = 3, fontface = 'bold', color = '#3C3C3C',
                  min.segment.length = 0) +
  annotate("text", x = 2000, y = max(yearly_range$MeanRange) - 0.2,
           label = paste0("Trend: ", slope_range, "\u00B0C per decade"),
           size = 3.5, fontface = 'bold', color = '#CD3333') +
  theme_blr() +
  labs(
    title = "Summer Diurnal Range Is Shrinking",
    subtitle = "The gap between daily high and low is narrowing as nights warm faster. A signature of urbanisation.",
    x = "", y = "Daily high minus daily low (\u00B0C)",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```


## 2024: the outlier year

2024 stands out in every metric. Let's see what made it special.

```{r year-2024, fig.width=12, fig.height=6}
# Daily highs in 2024 vs historical envelope
summer_daily %>%
  mutate(DOY = yday(Date)) %>%
  filter(DOY >= 60, DOY <= 152) ->  # Mar 1 to May 31
  summer_doy

summer_doy %>%
  filter(Year < 2024) %>%
  summarise(
    P10 = quantile(High, 0.10),
    P90 = quantile(High, 0.90),
    Median = median(High),
    .by = DOY
  ) ->
  hist_envelope

summer_doy %>%
  filter(Year == 2024) %>%
  select(DOY, High2024 = High) ->
  y2024

hist_envelope %>%
  left_join(y2024, by = "DOY") %>%
  mutate(Date = as.Date(DOY - 1, origin = "2024-01-01")) ->
  envelope_data

envelope_data %>%
  ggplot(aes(x = Date)) +
  geom_ribbon(aes(ymin = P10, ymax = P90), fill = "#9c9280", alpha = 0.3) +
  geom_line(aes(y = Median), color = "#9c9280", linewidth = 0.6) +
  geom_line(aes(y = High2024), color = "#CD3333", linewidth = 0.8) +
  geom_point(
    data = envelope_data %>% filter(!is.na(High2024) & High2024 > P90),
    aes(y = High2024), color = "#CD3333", size = 1.5
  ) +
  # Inline legend
  annotate("segment", x = as.Date("2024-03-05"), xend = as.Date("2024-03-15"),
           y = 38.5, yend = 38.5, color = "#CD3333", linewidth = 1) +
  annotate("text", x = as.Date("2024-03-16"), y = 38.5,
           label = "2024", hjust = 0, size = 3, fontface = 'bold', color = "#CD3333") +
  annotate("segment", x = as.Date("2024-03-05"), xend = as.Date("2024-03-15"),
           y = 37.5, yend = 37.5, color = "#9c9280", linewidth = 0.8) +
  annotate("text", x = as.Date("2024-03-16"), y = 37.5,
           label = "Median (1981-2023)", hjust = 0, size = 3, fontface = 'bold', color = "#9c9280") +
  annotate("rect", xmin = as.Date("2024-03-04"), xmax = as.Date("2024-03-15"),
           ymin = 36.3, ymax = 36.9, fill = "#9c9280", alpha = 0.3) +
  annotate("text", x = as.Date("2024-03-16"), y = 36.6,
           label = "10th-90th percentile", hjust = 0, size = 3, fontface = 'bold', color = "#9c9280") +
  scale_x_date(date_labels = "%b %d", breaks = seq(as.Date("2024-03-01"), as.Date("2024-05-31"), by = "2 weeks")) +
  theme_blr() +
  labs(
    title = "2024: Off the Charts",
    subtitle = "Daily highs in 2024 vs the 1981-2023 historical range. Almost the entire April-May period was above the 90th percentile.",
    x = "", y = "Daily high (\u00B0C)",
    caption = "Data: ERA5 via Oikolab"
  )
```


## How many hours above 30 C per summer day?

Another way to measure heat: not just the peak, but how long it stays hot.

```{r hours-above-30, fig.width=10, fig.height=5}
summer %>%
  mutate(Above30 = Temp >= 30) %>%
  summarise(HoursAbove30 = sum(Above30), .by = c(Date, Year)) %>%
  summarise(MeanHoursAbove30 = mean(HoursAbove30), .by = Year) ->
  hours_above

lm_hours <- lm(MeanHoursAbove30 ~ Year, data = hours_above)
slope_hours <- round(coef(lm_hours)[2] * 10, 1)

hours_above %>%
  mutate(
    Label = case_when(
      MeanHoursAbove30 == max(MeanHoursAbove30) | MeanHoursAbove30 == min(MeanHoursAbove30) ~
        paste0(Year, ": ", round(MeanHoursAbove30, 1), "h"),
      TRUE ~ ""
    )
  ) %>%
  ggplot(aes(x = Year, y = MeanHoursAbove30)) +
  geom_point(color = "#5f3946", size = 2) +
  geom_smooth(method = 'lm', se = FALSE, color = "#CD3333", linewidth = 0.7, linetype = 'dashed') +
  geom_text_repel(aes(label = Label), size = 3, fontface = 'bold', color = '#3C3C3C',
                  min.segment.length = 0) +
  annotate("text", x = 2000, y = max(hours_above$MeanHoursAbove30) - 0.2,
           label = paste0("Trend: +", slope_hours, " hours per decade"),
           size = 3.2, fontface = 'bold', color = '#CD3333') +
  theme_blr() +
  labs(
    title = "Hours Above 30\u00B0C Per Day in Summer",
    subtitle = "The number of uncomfortable hours is growing. Not just the peak, but the duration.",
    x = "", y = "Mean hours above 30\u00B0C per day",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  )
```


## The night question: warm nights are the new normal

Below 20 C on a summer night was common in the 1980s. Now it's rare.

```{r cool-nights-disappearing, fig.width=10, fig.height=5}
summer_daily %>%
  summarise(
    NightsBelow20 = sum(Low < 20),
    NightsAbove22 = sum(Low > 22),
    TotalDays = n(),
    .by = Year
  ) %>%
  mutate(
    PctBelow20 = NightsBelow20 / TotalDays * 100,
    PctAbove22 = NightsAbove22 / TotalDays * 100
  ) ->
  night_counts

p_cool <- night_counts %>%
  ggplot(aes(x = Year, y = PctBelow20)) +
  geom_col(fill = "#2255AA", width = 0.7) +
  geom_smooth(method = 'loess', se = FALSE, color = '#3C3C3C', linewidth = 0.5, span = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = function(x) paste0(x, "%")) +
  theme_blr() +
  labs(title = "Cool Summer Nights (Low < 20\u00B0C)", x = "", y = "")

p_warm <- night_counts %>%
  ggplot(aes(x = Year, y = PctAbove22)) +
  geom_col(fill = "#CD3333", width = 0.7) +
  geom_smooth(method = 'loess', se = FALSE, color = '#3C3C3C', linewidth = 0.5, span = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = function(x) paste0(x, "%")) +
  theme_blr() +
  labs(title = "Warm Summer Nights (Low > 22\u00B0C)", x = "", y = "")

p_cool + p_warm +
  plot_annotation(
    subtitle = "Cool nights almost gone; warm nights now dominate. This is the change people feel.",
    caption = "Data: ERA5 via Oikolab, 1981-2025"
  ) &
  theme(
    plot.subtitle = element_text(size = 9, color = '#5f3946'),
    panel.background = element_rect(fill = "#e5e1d8"),
    plot.background = element_rect(fill = "#e5e1d8", linewidth = 0)
  )
```


## Summary

```{r summary}
cat("ARE BANGALORE SUMMERS GETTING HOTTER?\n")
cat("=====================================\n\n")
cat("The short answer: yes, but not in the way Twitter thinks.\n\n")
cat("1. AFTERNOON HIGHS: barely moving\n")
cat(paste0("   +", slope_mean, " C/decade. A typical summer afternoon in 2025 is\n"))
cat(paste0("   only ~0.5 C warmer than in 1985.\n\n"))
cat("2. EXTREME DAYS: spiking\n")
cat("   Days above 35 C are getting more frequent and clustering\n")
cat("   into longer streaks. 2024 was unprecedented.\n\n")
cat("3. NIGHTS: this is where the action is\n")
cat(paste0("   +", slope_low, " C/decade. Summer nights have warmed ~1.5 C since 1981.\n"))
cat("   Cool nights below 20 C have nearly disappeared.\n\n")
cat("4. DURATION: not just hotter, longer\n")
cat(paste0("   +", slope_hours, " hours above 30 C per day, per decade.\n\n"))
cat("5. DIURNAL RANGE: shrinking\n")
cat(paste0("   ", slope_range, " C/decade. Nights warming faster than days\n"))
cat("   is a classic urban heat island signature.\n\n")
cat("What people feel as 'hotter summers' is likely:\n")
cat("   - More extreme heat days (spiky, not gradual)\n")
cat("   - Warmer nights with no relief\n")
cat("   - Less about 3 PM temperature, more about 3 AM\n")
```
