---
title: "Weather notebook"
---

```{r}
library(tidyverse)
library(lubridate)
load('~/Documents/work/data work/bangalore/weather/data/bangaloreRainfall.RData')
load('~/Documents/work/data work/bangalore/weather/data/bangaloreTemperature.RData')
load('~/Documents/work/data work/bangalore/weather/data/bangaloreWind.RData')
```

```{r}
blrWind
```

  Look at the avg wind speed by day of the year 
  
```{r}
blrWind %>%
  mutate(
    Day = day(DT),
    Month = month(DT),
    Wind = as.numeric(Wind)
  ) %>%
  summarise(
    WindSpeed = mean(Wind, na.rm=T),
    .by = c(Day, Month)
  ) %>%
  mutate(
    Year = year(Sys.Date()),
    Dt = make_date(Year, Month, Day)
  ) %>%
  ggplot(aes(x = Dt, y = WindSpeed)) + geom_point() + geom_line() + scale_x_date(date_breaks = '1 month', date_labels = '%b')
```
  
  Comapre to this year
  
```{r}
blrWind %>%
  filter(DT < Sys.Date()) %>%
  mutate(
    Year = year(Sys.Date()),
    Day = day(DT),
    Month = month(DT),
    Wind = as.numeric(Wind)
  ) %>%
  summarise(
    WindSpeed = mean(Wind[year(DT) < Year], na.rm=T),
    WindSpeedThisYear = mean(Wind[year(DT) == Year], na.rm=T),
    .by = c(Day, Month, Year)
  ) %>%
  mutate(
    DT = make_date(Year, Month, Day)
  )  %>%
  ggplot(aes(x = DT, y = WindSpeed)) + geom_point() + geom_line() + scale_x_date(date_breaks = '1 month', date_labels = '%b') + geom_line(aes(y = WindSpeedThisYear), col = 'red')

```
  

Yearly average temperatures by day overlaid with this year 

```{r}
blrTemp %>%
  filter(DT < Sys.Date()) %>%
  mutate(
    Year = year(DT),
    Day = day(DT),
    Month = month(DT),
    Temp = as.numeric(Temp)
  ) %>%
  summarise(
    AvgTemp = mean(Temp, na.rm=T),
    .by = c(Day, Month, Year)
  ) %>%
  mutate(
    Year1 = year(Sys.Date()),
    DT = make_date(Year1, Month, Day),
    ThisYear = Year == Year1
  )  %>%
  ggplot(aes(x = DT, y = AvgTemp, group = Year)) + geom_line(aes(col =  ThisYear), linewidth = 0.1) + scale_colour_manual(values = c('grey', 'black')) + theme_minimal() +  scale_x_date(date_breaks = '1 month', date_labels = '%b') + theme(legend.position = 'none')
```

What is the likelihood of raining in the evenings in May? And how does this vary by month? 


```{r}
blrRain %>%
  filter(month(DT) == 5) %>% 
  filter(hour(DT) >= 15 & hour(DT) <= 18) %>% 
  mutate(Hour = hour(DT)) %>%
  summarise(
    RainProb = mean(Rain > 1),
    .by = Hour
  )
```

And now by how the month goes as well 

```{r}
blrRain %>%
  filter(month(DT) == 5) %>% 
  filter(hour(DT) >= 15 & hour(DT) <= 18) %>% 
  mutate(Hour = hour(DT), Day = day(DT), Rain = as.numeric(Rain)) %>%
  summarise(
    RainProb = mean(Rain > 1),
    .by = c(Hour, Day)
  ) %>% 
  ggplot(aes(x = Day, y= RainProb)) + geom_point() + geom_line() + facet_wrap(~Hour)
```

What matters is the likelihood that it will rain sometime in this window - not at a specific time


```{r}
blrRain %>%
  filter(month(DT) == 5) %>% 
  filter(hour(DT) >= 15 & hour(DT) < 18) %>% 
  mutate(Year = year(DT), Day = day(DT), Rain =as.numeric(Rain)) %>%
  summarise(
    Rain = sum(Rain), .by = c(Year, Day)
  ) %>%
  summarise(
    RainProb = mean(Rain > 1),
    .by = Day
  ) %>% 
  ggplot(aes(x = Day, y= RainProb)) + geom_point() + geom_line() + theme_minimal() + scale_y_continuous(lim =c(0,1)) + geom_smooth() 
```

What if i put limit to 2mm? or 3mm? or 5mm? 

```{r}
blrRain %>%
  filter(month(DT) == 5) %>% 
  filter(hour(DT) >= 15 & hour(DT) < 18) %>% 
  mutate(Year = year(DT), Day = day(DT), Rain = as.numeric(Rain)) %>%
  summarise(
    Rain = sum(Rain), .by = c(Year, Day)
  ) %>%
  crossing(
    Limit = 2:5
  ) %>%
  summarise(
    RainProb = mean(Rain > Limit),
    .by = c(Day, Limit)
  ) %>% 
  ggplot(aes(x = Day, y= RainProb)) + geom_point() + geom_line() + theme_minimal() + scale_y_continuous(lim =c(0,1)) + geom_smooth() + facet_wrap(~Limit)
```

Just summarise

```{r}
blrRain %>%
  filter(month(DT) == 5) %>% 
  filter(hour(DT) >= 15 & hour(DT) < 18) %>% 
  mutate(Year = year(DT), Day = day(DT), Rain = as.numeric(Rain)) %>%
  summarise(
    Rain = sum(Rain), .by = c(Year, Day)
  ) %>%
  crossing(
    Limit = 1:5
  ) %>%
  summarise(
    RainProb = mean(Rain > Limit),
    .by = c(Limit)
  ) 
```

Look at 7pm temperatures through the year. Compare this year to average

```{r}
blrTemp %>%
  filter(DT < Sys.Date()) %>%
  filter(hour(DT) == 19) %>%
  mutate(
    Year = year(DT),
    Day = day(DT),
    Month = month(DT),
    Temp = as.numeric(Temp)
  ) %>%
  mutate(
    AvgTemp = median(Temp, na.rm=T),
    ThisTemp = Temp[Year == max(Year)][1], 
    .by = c(Day, Month)
  ) %>%
  mutate(
    Year1 = year(Sys.Date()),
    DT = make_date(Year1, Month, Day),
    ThisYear = Year == Year1,
    ThisTemp = ifelse(Year == Year1, ThisTemp, NA)
  )  %>%
  ggplot(aes(x = DT, y = Temp, group = Year)) + geom_line(linewidth = 0.05) + geom_line(aes(y = AvgTemp), linewidth = 0.5, col = "#490000", lty = 2) + geom_line(aes(y = ThisTemp), linewidth = 0.7, col = "#490000")  + theme_minimal() +  scale_x_date(date_breaks = '1 month', date_labels = '%b') + theme(legend.position = 'none') +  ggtitle("Bangalore 7 pm temperature") + theme_tufte() + 
  theme(panel.grid = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_line(colour = 'black', linewidth = 0.1), axis.text.x = element_text(face = 'bold', size = 7), axis.line.y = element_line(colour = 'black', linewidth = 0.2),  panel.background = element_rect(fill = "#eae4db", linewidth = 0), plot.background = element_rect(fill = "#eae4db") )
```



