---
title: "April 30 and May 2 rain data check"
question: "Are the heavy Bangalore rains around April 30 and May 2 missing from the local weather data?"
trigger_prompt: "can you check the data once again? we had fairly heavy rain in bangalore on apr30 and may2. but they don't seem to be recorded here at all"
slug: "apr30-may2-rain-data-check"
date: "2026-05-28"
analyst: "Codex"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 6,
  dpi = 150
)

library(tidyverse)
library(lubridate)
library(scales)

root_dir <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
chart_dir <- file.path(root_dir, "docs", "assets", "analysis")
dir.create(chart_dir, recursive = TRUE, showWarnings = FALSE)

load(file.path(root_dir, "data", "bangaloreTemperature.RData"))
load(file.path(root_dir, "data", "bangaloreRainfall.RData"))
```

## Question

`r rmarkdown::metadata$question`

## Data Prep

```{r data-prep}
rain_hourly <- blrRain %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    day = day(DT),
    hour = hour(DT),
    source = Source,
    rain_mm = as.numeric(Rain)
  ) %>%
  filter(!is.na(rain_mm))

rain_daily <- rain_hourly %>%
  summarise(
    hours = n(),
    rain_mm = sum(rain_mm, na.rm = TRUE),
    max_hour_mm = max(rain_mm, na.rm = TRUE),
    .by = c(date, source)
  )

rain_hourly %>%
  summarise(
    first_dt = min(DT),
    last_dt = max(DT),
    rows = n(),
    sources = paste(sort(unique(source)), collapse = ", ")
  )
```

## Local Check

```{r local-daily-window}
rain_daily %>%
  filter(date >= as_date("2026-04-25"), date <= as_date("2026-05-05")) %>%
  arrange(date, source)
```

```{r target-hourly}
rain_hourly %>%
  filter(date %in% as_date(c("2026-04-30", "2026-05-02"))) %>%
  select(DT, date, source, rain_mm) %>%
  arrange(DT)
```

```{r exported-csv-check}
exported_daily <- read_csv(
  file.path(root_dir, "docs", "data", "bangalore_daily_weather.csv"),
  show_col_types = FALSE
) %>%
  mutate(DT = as_date(DT))

exported_daily %>%
  filter(DT >= as_date("2026-04-25"), DT <= as_date("2026-05-05"))
```

The current local data does not have both days as zero. April 30 is now present as
8.85 mm in the ERA5/Oikolab hourly file and in the exported CSV. May 2 is still
0.00 mm.

## External Context

This is not just a user-memory mismatch. Multiple public reports describe a major
Bengaluru rain event on April 29/30, with IMD city-observatory totals around
110-111 mm and heavy waterlogging. Search results also surfaced a May 2 report of
intense rainfall in central/east localities such as MG Road, Shivajinagar,
Indiranagar, and Commercial Street.

Useful reference URLs checked:

- https://indianexpress.com/article/cities/bangalore/bengaluru-rainfall-record-april-111mm-deaths-damage-10663611/
- https://newsable.asianetnews.com/karnataka-news/bengaluru-rains-imd-records-111-mm-rainfall-highest-ever-april-orange-alert-issued-severe-hailstorm-articleshow-unfr05v
- https://newskarnataka.com/bengaluru/sudden-cloudburst-like-rains-overwhelm-central-bengaluru/02052026/

## Checks

```{r historical-context}
target_days <- rain_daily %>%
  filter(date %in% as_date(c("2026-04-30", "2026-05-02"))) %>%
  select(date, source, rain_mm, max_hour_mm)

calendar_day_context <- rain_daily %>%
  mutate(md = format(date, "%m-%d")) %>%
  filter(year(date) < 2026, md %in% c("04-30", "05-02")) %>%
  summarise(
    years = n(),
    median_mm = median(rain_mm),
    p90_mm = quantile(rain_mm, 0.9),
    max_mm = max(rain_mm),
    .by = md
  )

target_days
calendar_day_context
```

April 30's 8.85 mm is high relative to this grid's own April 30 history, but it is
far below the reported city-observatory event. May 2 is in the lower tail for this
grid and contradicts reports of localized heavy rain in parts of the city.

The likely problem is spatial representativeness. The project uses a point/grid
series for Bangalore from Oikolab/ERA5/GFS. Short pre-monsoon thunderstorms can be
extremely localized. A city event can be severe in central, south, or east
Bengaluru while a gridded point or an airport-adjacent station misses most of the
rain. This also explains why a later May 21 report noted 44.2 mm at the city
observatory but only 3.8 mm near HAL.

## Verdict

Insightful: no

April 30 is no longer absent in the current local export: it is recorded as
8.85 mm. But that is still a severe undercount against public IMD-style reports
near 110-111 mm for the city event. May 2 remains absent from the local rainfall
series at 0.00 mm, despite reports of intense localized rain in central/east
Bengaluru.

This should not be published as a blog post. It is a data QA finding: the rain
series is not reliable for localized convective storm totals, and the site should
carry a stronger caveat or switch/augment rainfall with an observed city rain-gauge
source if those dates matter.

Charts:

- none

Blog post:

- not published
