---
title: "Bangalore morning rain between 8-10am"
question: "When does it rain in Bangalore in the morning 8-10 am? do we ever require an umbrella then?"
trigger_prompt: "when does it rain in Bangalore in the morning 8-10 am? do we ever require an umbrella then?"
slug: "bangalore-morning-rain-8to10"
date: "`r Sys.Date()`"
analyst: "Codex"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 7,
  dpi = 160
)

library(tidyverse)
library(lubridate)
library(scales)

root_dir <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
chart_dir <- file.path(root_dir, "docs", "assets", "analysis")
dir.create(chart_dir, recursive = TRUE, showWarnings = FALSE)

load(file.path(root_dir, "data", "bangaloreRainfall.RData"))

question_theme <- function(base_size = 11) {
  theme_minimal(base_size = base_size, base_family = "Helvetica") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
      axis.title = element_blank(),
      axis.text = element_text(color = "#3C3C3C", face = "bold"),
      plot.title = element_text(color = "#3C3C3C", face = "bold"),
      plot.subtitle = element_text(color = "#666666"),
      plot.caption = element_text(color = "#666666", size = rel(0.85)),
      legend.title = element_blank()
    )
}

fmt_pct <- label_percent(accuracy = 1)
```

## Question

`r rmarkdown::metadata$question`

## Analysis contract

- **Unit of analysis:** one observed hourly record.
- **Main denominator:** all observed hours in 1981-2025, or all 8-10am hours when looking at the commute window.
- **Main metric:** share of hours with rain > 0.1 mm.
- **Sensitivity check:** rain > 0 mm gives the same broad timing.
- **Comparison:** month-by-month timing, with a focus on the 8-10am window.
- **Umbrella proxy:** any rain in the 8-10am window on a given day.

## Data prep

```{r data-prep}
rain_hourly <- blrRain %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    hour = hour(DT),
    rain_mm = as.numeric(Rain)
  ) %>%
  filter(!is.na(rain_mm), year >= 1981, year <= 2025)

rain_summary <- rain_hourly %>%
  summarise(
    first_dt = min(DT),
    last_dt = max(DT),
    rows = n(),
    days = n_distinct(date),
    years = n_distinct(year)
  )

rain_summary
```

## What the archive says

The 8-10am window is *not* Bangalore's wettest time of day. Rain in that window is mostly a monsoon and post-monsoon story: it picks up sharply from June, peaks in July-August, stays fairly active into October-November, and is rare in the driest months.

Within the morning itself, 10am is the rainiest of the three hours. Across the full 1981-2025 archive, the share of rainy hours is about 0.9% at 8am, 1.5% at 9am, and 2.5% at 10am in January; it jumps to about 19.5%, 21.9%, and 31.6% in July.

```{r facts-table}
morning_hours <- rain_hourly %>%
  filter(hour %in% 8:10) %>%
  summarise(
    rainy_share = mean(rain_mm > 0.1),
    mean_mm = mean(rain_mm),
    median_mm = median(rain_mm),
    p90_mm = as.numeric(quantile(rain_mm, 0.9)),
    .by = c(month, hour)
  ) %>%
  mutate(
    month_name = month.abb[month],
    rainy_share = round(rainy_share, 3),
    mean_mm = round(mean_mm, 3),
    median_mm = round(median_mm, 3),
    p90_mm = round(p90_mm, 3)
  ) %>%
  arrange(month, hour)

morning_hours
```

```{r window-days}
morning_window <- rain_hourly %>%
  filter(hour %in% 8:10) %>%
  summarise(
    window_mm = sum(rain_mm),
    rainy_any = any(rain_mm > 0.1),
    rainy_hours = sum(rain_mm > 0.1),
    .by = c(date, year, month)
  )

window_by_month <- morning_window %>%
  summarise(
    share = mean(rainy_any),
    mean_mm = mean(window_mm),
    p90_mm = as.numeric(quantile(window_mm, 0.9)),
    .by = month
  ) %>%
  mutate(
    month_name = month.abb[month],
    share = round(share, 3),
    mean_mm = round(mean_mm, 3),
    p90_mm = round(p90_mm, 3)
  ) %>%
  arrange(month)

window_by_month
```

A useful commute-window read is: a day gets at least one rainy 8-10am hour about **17%** of the time overall, but that rises to **25-39%** in June-August and stays elevated into October-November.

```{r seasonal-check}
seasonal_window <- morning_window %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "DJF",
    month %in% 3:5 ~ "MAM",
    month %in% 6:9 ~ "JJA",
    TRUE ~ "OND"
  )) %>%
  summarise(
    share = mean(rainy_any),
    mean_mm = mean(window_mm),
    .by = season
  ) %>%
  mutate(
    share = round(share, 3),
    mean_mm = round(mean_mm, 3)
  ) %>%
  arrange(match(season, c("DJF", "MAM", "JJA", "OND")))

seasonal_window
```

## Chart 1: where 8-10am fits in the daily rain clock

```{r chart-heatmap, fig.width=12, fig.height=7}
heatmap_data <- rain_hourly %>%
  summarise(
    rainy_share = mean(rain_mm > 0.1),
    .by = c(month, hour)
  ) %>%
  mutate(month_name = factor(month.abb[month], levels = rev(month.abb)))

p_heat <- ggplot(heatmap_data, aes(x = hour, y = month_name, fill = rainy_share)) +
  geom_tile(color = "white", linewidth = 0.25) +
  annotate(
    "rect",
    xmin = 7.5,
    xmax = 10.5,
    ymin = 0.5,
    ymax = 12.5,
    fill = NA,
    color = "#CD3333",
    linewidth = 0.7,
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = seq(0, 23, 2), labels = paste0(seq(0, 23, 2), ":00"), limits = c(0, 23)) +
  scale_fill_gradient(low = "#f3efe8", high = "#228B22", labels = fmt_pct) +
  labs(
    title = "Rainy hours cluster in monsoon and post-monsoon, not in the 8-10am commute band",
    subtitle = "Share of hours with >0.1 mm rain, 1981-2025. Red box marks 8-10am.",
    caption = "Data: ERA5 via Oikolab"
  ) +
  question_theme() +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = "bottom",
    legend.key.width = unit(1.8, "cm")
  )

p_heat

chart_heatmap_path <- file.path(chart_dir, "bangalore_morning_rain_heatmap.png")
ggsave(chart_heatmap_path, p_heat, width = 12, height = 7, dpi = 180, bg = "white")
```

## Chart 2: how often the 8-10am window needs an umbrella

```{r chart-window, fig.width=12, fig.height=6}
p_window <- ggplot(window_by_month, aes(x = month, y = share)) +
  geom_col(fill = "#5f3946", width = 0.7) +
  geom_text(
    aes(label = paste0(month_name, "\n", round(share * 100), "%")),
    vjust = -0.35,
    size = 3.2,
    fontface = "bold",
    color = "#3C3C3C",
    lineheight = 0.9
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = fmt_pct, limits = c(0, 0.45), expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "An umbrella is mainly a monsoon/post-monsoon morning concern",
    subtitle = "Share of days with at least one rainy hour between 8am and 10am",
    caption = "Rainy day = at least one hour in the window above 0.1 mm"
  ) +
  question_theme()

p_window

chart_window_path <- file.path(chart_dir, "bangalore_morning_rain_window_share.png")
ggsave(chart_window_path, p_window, width = 12, height = 6, dpi = 180, bg = "white")
```

## Checks

```{r checks}
rain_hourly %>%
  filter(hour %in% 8:10) %>%
  summarise(
    share_gt_0 = mean(rain_mm > 0),
    share_gt_0_1 = mean(rain_mm > 0.1),
    share_gt_1 = mean(rain_mm > 1),
    .by = hour
  )
```

The broad timing survives a looser `> 0 mm` threshold: 10am is still the wettest morning hour, and the same wet-season shape remains. The exact percentages move a little, but the story does not.

## Verdict

Insightful: no

This is a useful commuting answer, but it is not novel enough to publish as a standalone post. The pattern is straightforward once you look at the hourly archive: 8-10am rain is mostly a June-November phenomenon, with 10am the rainiest of the three hours, and the strongest umbrella need in July-August.

It sharpens the intuition a bit, but it does not clear the publication bar because the effect is seasonal and descriptive rather than surprising. Keep it as a notebook answer, not a blog post.

Charts:

- `docs/assets/analysis/bangalore_morning_rain_heatmap.png`
- `docs/assets/analysis/bangalore_morning_rain_window_share.png`

Blog post:

- not published
- if this ever becomes publishable, create `docs/blog/bangalore-morning-rain-8to10.html`
- include `This post is AI-written.` at the top of the article body
- end with an `Original question` section containing `r rmarkdown::metadata$trigger_prompt`
- add the post to the top of `docs/blog/index.html` and `docs/blog/feed.xml`
- add `blog_url: "blog/bangalore-morning-rain-8to10.html"` to the top of `analysis_questions/analyses.yml`
