---
title: "Bangalore rain has a 4pm habit"
question: "Does it rain at specific hours of the day during monsoons in Bangalore, versus rest of the year?"
trigger_prompt: "one weather question - does it rain at specific hours of the day during monsoons in bangalore, versus rest of the year?"
slug: "bangalore-monsoon-rain-hours"
date: "2026-06-25"
analyst: "Codex"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 6,
  dpi = 150
)

library(tidyverse)
library(lubridate)
library(scales)

root_dir <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
chart_dir <- file.path(root_dir, "docs", "assets", "analysis")
dir.create(chart_dir, recursive = TRUE, showWarnings = FALSE)

load(file.path(root_dir, "data", "bangaloreRainfall.RData"))

rain_hourly <- blrRain %>%
  filter(str_to_lower(Source) == "era5") %>%
  transmute(
    DT = as.POSIXct(DT, tz = "Asia/Kolkata"),
    rain_mm = as.numeric(Rain)
  ) %>%
  mutate(
    month = month(DT),
    hour = hour(DT),
    season = if_else(month %in% c(6, 7, 8, 9), "Monsoon", "Rest"),
    rainy = rain_mm > 0.1,
    rainy_any = rain_mm > 0
  )

summary_by_hour <- rain_hourly %>%
  summarise(
    rainy_pct = mean(rainy) * 100,
    mean_rain = mean(rain_mm),
    .by = c(season, hour)
  ) %>%
  mutate(season = factor(season, levels = c("Rest", "Monsoon")))

period_summary <- rain_hourly %>%
  mutate(period = case_when(
    hour %in% 0:5 ~ "Night (0-5)",
    hour %in% 6:11 ~ "Morning (6-11)",
    hour %in% 12:17 ~ "Afternoon (12-17)",
    TRUE ~ "Evening (18-23)"
  )) %>%
  summarise(
    rainy_pct = mean(rainy) * 100,
    mean_rain = mean(rain_mm),
    total_rain = sum(rain_mm),
    .by = c(season, period)
  ) %>%
  mutate(
    season = factor(season, levels = c("Rest", "Monsoon")),
    period = factor(period, levels = c("Night (0-5)", "Morning (6-11)", "Afternoon (12-17)", "Evening (18-23)"))
  )

peak_hours <- summary_by_hour %>%
  group_by(season) %>%
  slice_max(rainy_pct, n = 1, with_ties = FALSE) %>%
  ungroup()

threshold_check <- rain_hourly %>%
  summarise(rainy_any_pct = mean(rainy_any) * 100, .by = c(season, hour)) %>%
  group_by(season) %>%
  slice_max(rainy_any_pct, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(season, peak_hour_any = hour)
```

## Question

`r rmarkdown::metadata$question`

## What the archive says

The rain clock is real in both seasons, but it is much sharper in monsoon. The wettest hour is the same in both cases — **4pm** — yet the monsoon peak is more than twice as rainy as the rest of the year. In the monsoon window, about **55%** of 4pm hours are rainy; outside monsoon, it is about **21%**.

That pattern is clearest in the afternoon. Monsoon afternoons (12-5pm) are rainy about **39.5%** of the time, versus **15.2%** in the rest of the year. Nights are much drier in both cases.

```{r peak-table}
peak_hours %>%
  mutate(
    rainy_pct = round(rainy_pct, 1),
    mean_rain = round(mean_rain, 3)
  )
```

```{r period-table}
period_summary %>%
  mutate(
    rainy_pct = round(rainy_pct, 1),
    mean_rain = round(mean_rain, 3),
    total_rain = round(total_rain, 1)
  )
```

## Chart

```{r rain-hour-chart}
chart_path <- file.path(chart_dir, "bangalore_monsoon_rain_diurnal.png")

plot_data <- summary_by_hour %>%
  mutate(season = fct_relevel(season, "Rest", "Monsoon"))

p <- ggplot(plot_data, aes(x = hour, y = rainy_pct, color = season)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.7) +
  geom_vline(xintercept = 16, linetype = "dashed", color = "#999999", linewidth = 0.4) +
  geom_text(
    data = peak_hours,
    aes(label = paste0(round(rainy_pct, 1), "% at ", hour, ":00")),
    nudge_y = 3,
    size = 3.3,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3), labels = paste0(seq(0, 23, 3), ":00")) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 60)) +
  scale_color_manual(values = c("Rest" = "#9c9280", "Monsoon" = "#228B22")) +
  labs(
    title = "Bangalore rain peaks in the late afternoon",
    subtitle = "Share of ERA5 hours with >0.1 mm rain, 1981-2026. The 4pm peak is much stronger during June-September.",
    x = NULL,
    y = "Rainy hours",
    caption = "Data: ERA5 via Oikolab"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
    axis.title = element_blank(),
    axis.text = element_text(color = "#3C3C3C", face = "bold"),
    plot.title = element_text(color = "#3C3C3C", face = "bold"),
    plot.subtitle = element_text(color = "#666666"),
    plot.caption = element_text(color = "#666666", size = rel(0.85)),
    legend.title = element_blank(),
    legend.position = "bottom"
  )

p

ggsave(chart_path, p, width = 12, height = 6, dpi = 180, bg = "white")
```

## Sensitivity check

The timing survives a looser rain definition too: if I count any rain at all (`> 0 mm`) instead of `> 0.1 mm`, the peak hour is still **4pm** in both seasons. The difference is mostly how often the afternoon gets wet, not a shift to some other part of the day.

```{r threshold-check}
threshold_check
```

## Verdict

Insightful: yes

The answer is a clear yes: Bangalore rain has a strong late-afternoon clock, and the monsoon version of that clock is much stronger than the rest of the year. That is enough to publish because it sharpens a plausible intuition with a simple, stable pattern across 45 years of hourly data. The story is not just "it rains more in monsoon"; it is "the wettest hours are concentrated around 4pm, and monsoon amplifies that concentration."

Charts:

- `docs/assets/analysis/bangalore_monsoon_rain_diurnal.png`

Blog post:

- publish `docs/blog/bangalore-monsoon-rain-hours.html`
- add to `docs/blog/index.html`
- add to `docs/blog/feed.xml`
- add `blog_url: "blog/bangalore-monsoon-rain-hours.html"` to `analysis_questions/analyses.yml`
