---
title: "Have Bangalore summers become hotter?"
question: "Have Bangalore summers become hotter over the years?"
slug: "bangalore-summers-hotter"
date: "2026-05-23"
analyst: "Codex"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 7,
  dpi = 160
)

library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)

root_dir <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
chart_dir <- file.path(root_dir, "docs", "assets", "analysis")
dir.create(chart_dir, recursive = TRUE, showWarnings = FALSE)

load(file.path(root_dir, "data", "bangaloreTemperature.RData"))

theme_question <- function(base_size = 11) {
  theme_minimal(base_size = base_size, base_family = "Helvetica") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
      axis.title = element_blank(),
      axis.text = element_text(color = "#3C3C3C", face = "bold"),
      plot.title = element_text(color = "#3C3C3C", face = "bold"),
      plot.subtitle = element_text(color = "#666666"),
      plot.caption = element_text(color = "#666666", size = rel(0.85)),
      strip.text = element_text(color = "#3C3C3C", face = "bold"),
      legend.position = "none"
    )
}

slope_per_decade <- function(data, y) {
  coef(lm(reformulate("year", y), data = data))[["year"]] * 10
}

label_slope <- function(x) {
  paste0(ifelse(x >= 0, "+", ""), number(x, accuracy = 0.01), " deg C/decade")
}
```

## Question

Have Bangalore summers become hotter over the years?

I treat summer as March-May, and use complete summers from 1981-2025. The local data
also contains part of summer 2026, but May 2026 is not complete as of this analysis, so
2026 is excluded from the trend estimates.

## Data Prep

```{r data-prep}
temp_hourly <- blrTemp %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    hour = hour(DT),
    temp_c = as.numeric(Temp)
  ) %>%
  filter(!is.na(temp_c), year >= 1981, year <= 2025)

summer_hourly <- temp_hourly %>%
  filter(month %in% 3:5)

summer_daily <- summer_hourly %>%
  summarise(
    tmax = max(temp_c, na.rm = TRUE),
    tmin = min(temp_c, na.rm = TRUE),
    tmean = mean(temp_c, na.rm = TRUE),
    .by = c(date, year)
  ) %>%
  filter(is.finite(tmax), is.finite(tmin), is.finite(tmean))

coverage <- summer_daily %>%
  summarise(days = n(), .by = year) %>%
  summarise(
    first_year = min(year),
    last_year = max(year),
    complete_summers = sum(days >= 92),
    min_days = min(days),
    max_days = max(days)
  )

coverage
```

## Yearly Summer Metrics

```{r yearly-metrics}
yearly <- summer_daily %>%
  summarise(
    mean_high = mean(tmax),
    mean_low = mean(tmin),
    mean_temp = mean(tmean),
    mean_range = mean(tmax - tmin),
    days_35 = sum(tmax >= 35),
    days_36 = sum(tmax >= 36),
    max_high = max(tmax),
    .by = year
  )

trend_summary <- tibble(
  metric = c("Mean daily high", "Mean daily low", "Mean temperature", "Daily high-low range", "Days at or above 35 deg C"),
  trend_per_decade = c(
    slope_per_decade(yearly, "mean_high"),
    slope_per_decade(yearly, "mean_low"),
    slope_per_decade(yearly, "mean_temp"),
    slope_per_decade(yearly, "mean_range"),
    coef(lm(days_35 ~ year, data = yearly))[["year"]] * 10
  )
)

trend_summary
```

The main pattern is asymmetric. The average summer daily high has not moved much over
the full 1981-2025 archive, while the average summer daily low has risen clearly.
That means Bangalore's summer heat is showing up more in nights and persistence than
in the average afternoon maximum.

## Main Chart

```{r main-chart, fig.width=12, fig.height=9}
high_low <- yearly %>%
  select(year, `Daily high` = mean_high, `Daily low` = mean_low) %>%
  pivot_longer(-year, names_to = "metric", values_to = "temp_c") %>%
  mutate(metric = factor(metric, levels = c("Daily high", "Daily low")))

end_labels <- high_low %>%
  group_by(metric) %>%
  filter(year == max(year)) %>%
  ungroup() %>%
  left_join(
    high_low %>%
      summarise(slope = slope_per_decade(cur_data(), "temp_c"), .by = metric),
    by = "metric"
  ) %>%
  mutate(label = paste0(metric, "\n", label_slope(slope)))

p_trend <- ggplot(high_low, aes(year, temp_c, color = metric)) +
  geom_point(size = 1.4, alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9, linetype = "dashed") +
  geom_text(
    data = end_labels,
    aes(label = label),
    hjust = 0,
    nudge_x = 1,
    size = 3.2,
    fontface = "bold",
    lineheight = 0.95
  ) +
  scale_color_manual(values = c("Daily high" = "#CD3333", "Daily low" = "#2255AA")) +
  scale_x_continuous(limits = c(1981, 2034), breaks = seq(1985, 2025, 10)) +
  scale_y_continuous(labels = label_number(suffix = " deg C")) +
  labs(
    title = "Summer highs are nearly flat; summer lows are rising",
    subtitle = "Mean daily maximum and minimum temperature in March-May, 1981-2025"
  ) +
  theme_question()

hot_labels <- yearly %>%
  filter(days_35 >= 10) %>%
  mutate(label = paste0(year, "\n", days_35, " days"))

p_hot <- ggplot(yearly, aes(year, days_35)) +
  geom_col(fill = "#5f3946", width = 0.72) +
  geom_smooth(method = "loess", se = FALSE, color = "#CD3333", linewidth = 0.8, span = 0.55) +
  geom_text(
    data = hot_labels,
    aes(label = label),
    vjust = -0.25,
    size = 2.8,
    fontface = "bold",
    color = "#3C3C3C",
    lineheight = 0.9
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(
    title = "Hot-day spikes are larger in recent decades",
    subtitle = "Number of March-May days with daily maximum at or above 35 deg C"
  ) +
  theme_question()

p_range <- ggplot(yearly, aes(year, mean_range)) +
  geom_point(color = "#005566", size = 1.5, alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE, color = "#005566", linewidth = 0.9, linetype = "dashed") +
  annotate(
    "text",
    x = 1984,
    y = max(yearly$mean_range) - 0.15,
    hjust = 0,
    label = paste0(
      "Daily range trend: ",
      label_slope(slope_per_decade(yearly, "mean_range"))
    ),
    color = "#005566",
    size = 3.2,
    fontface = "bold"
  ) +
  scale_y_continuous(labels = label_number(suffix = " deg C")) +
  labs(
    title = "The daily high-low range is shrinking",
    subtitle = "Summer nights are warming faster than summer afternoons"
  ) +
  theme_question()

main_chart <- (p_trend / p_hot / p_range) +
  plot_annotation(
    title = "Have Bangalore summers become hotter?",
    subtitle = "Yes, but the clearest change is warmer nights and bigger hot spells, not a large rise in the average afternoon high.",
    caption = "Hourly Oikolab/ERA5 archive for Bangalore. Summer = March-May. Complete summers only, 1981-2025.",
    theme = theme(
      plot.title = element_text(color = "#3C3C3C", face = "bold", size = 18),
      plot.subtitle = element_text(color = "#666666", size = 11),
      plot.caption = element_text(color = "#666666", size = 8)
    )
  )

main_chart

chart_path <- file.path(chart_dir, "bangalore_summer_warming.png")
ggsave(chart_path, main_chart, width = 12, height = 9, dpi = 180, bg = "white")
```

## Checks

```{r checks}
era_compare <- yearly %>%
  mutate(era = case_when(
    year <= 1995 ~ "1981-1995",
    year >= 2011 ~ "2011-2025",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(era)) %>%
  summarise(
    mean_high = mean(mean_high),
    mean_low = mean(mean_low),
    days_35 = mean(days_35),
    max_high = mean(max_high),
    .by = era
  )

era_compare

yearly %>%
  arrange(desc(days_35)) %>%
  select(year, mean_high, mean_low, days_35, days_36, max_high) %>%
  slice_head(n = 10)
```

The result is robust to looking at eras instead of a linear trend: recent summers have
warmer nights and more hot days on average, even though year-to-year weather still
dominates any single summer. 2024 is an extreme outlier, so the claim should not be
phrased as "every summer is hotter than before." It should be phrased as a shift in
the distribution: nights have moved up, the daily high-low range has narrowed, and
large hot spells are easier to get.

## Verdict

Insightful: yes

Bangalore summers have become hotter, but not mainly because the average afternoon
maximum has marched upward. From 1981-2025, summer daily highs are roughly flat, while
summer daily lows rise by about 0.22 deg C per decade. The better public answer is:
summer heat has become more persistent and more nocturnal, with recent decades also
showing larger spikes in days above 35 deg C.

Charts:

- `docs/assets/analysis/bangalore_summer_warming.png`
