---
title: "Has July ever been this hot before?"
question: "Has July ever been this hot before, and what has hot July meant for overall seasonal (Jul-Sep) rainfall?"
trigger_prompt: "has July ever been this hot before? and hwat has hot July meant for overall seasonal (jul-sep) rainfall?"
slug: "july-hot-rainfall"
date: "`r Sys.Date()`"
analyst: "Codex"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 6,
  dpi = 160
)

library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)

root_dir <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
chart_dir <- file.path(root_dir, "docs", "assets", "analysis")
dir.create(chart_dir, recursive = TRUE, showWarnings = FALSE)

load(file.path(root_dir, "data", "bangaloreTemperature.RData"))
load(file.path(root_dir, "data", "bangaloreRainfall.RData"))

analysis_date <- max(as_datetime(blrTemp$DT), na.rm = TRUE) |> as_date()
current_year <- year(analysis_date)
cutoff_day <- day(analysis_date)

theme_weather_question <- function(base_size = 11) {
  theme_minimal(base_size = base_size, base_family = "Helvetica") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
      axis.title = element_blank(),
      axis.text = element_text(color = "#3C3C3C", face = "bold"),
      plot.title = element_text(color = "#3C3C3C", face = "bold"),
      plot.subtitle = element_text(color = "#666666"),
      plot.caption = element_text(color = "#666666", size = rel(0.85)),
      legend.position = "none"
    )
}

fmt_rank <- function(x, n) paste0(x, " / ", n)
```

## Question

`r rmarkdown::metadata$question`

## Analysis frame

- *July heat* is measured two ways:
  1. July 1-15 mean daily maximum temperature, to compare the current month-to-date heat fairly against past Julys at the same point in the month.
  2. Full-month July mean daily maximum temperature, to test whether a hot July has historically meant a wetter or drier Jul-Sep season.
- *Seasonal rainfall* means total rain in July-September of the same year.
- Data source is the local ERA5 archive only.

## Data prep

```{r data-prep}
temp_hourly <- blrTemp %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    day = day(DT),
    temp_c = as.numeric(Temp)
  ) %>%
  filter(!is.na(temp_c))

rain_hourly <- blrRain %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    day = day(DT),
    rain_mm = as.numeric(Rain)
  ) %>%
  filter(!is.na(rain_mm))

daily_weather <- temp_hourly %>%
  summarise(
    tmin = min(temp_c, na.rm = TRUE),
    tmax = max(temp_c, na.rm = TRUE),
    tmean = mean(temp_c, na.rm = TRUE),
    .by = c(date, year, month, day)
  ) %>%
  left_join(
    rain_hourly %>%
      summarise(rain_mm = sum(rain_mm, na.rm = TRUE), .by = c(date, year, month, day)),
    by = c("date", "year", "month", "day")
  ) %>%
  filter(is.finite(tmin), is.finite(tmax), !is.na(rain_mm))

daily_weather %>%
  summarise(
    first_date = min(date),
    last_date = max(date),
    days = n(),
    years = n_distinct(year)
  )
```

## July so far: is this already an unusually hot July?

```{r july-ytd-metrics}
july_ytd <- daily_weather %>%
  filter(month == 7, day <= cutoff_day) %>%
  summarise(
    mean_tmax = mean(tmax),
    mean_tmin = mean(tmin),
    mean_tmean = mean(tmean),
    days_30 = sum(tmax >= 30),
    days_31 = sum(tmax >= 31),
    days_32 = sum(tmax >= 32),
    days_33 = sum(tmax >= 33),
    days_35 = sum(tmax >= 35),
    max_tmax = max(tmax),
    rain_total = sum(rain_mm),
    rainy_days = sum(rain_mm > 0.1),
    .by = year
  ) %>%
  mutate(
    rank_mean_tmax = min_rank(desc(mean_tmax)),
    rank_max_tmax = min_rank(desc(max_tmax)),
    rank_days_33 = min_rank(desc(days_33))
  )

current_ytd <- july_ytd %>% filter(year == current_year)
current_ytd

thresholds_ytd <- july_ytd %>%
  summarise(
    across(c(days_30, days_31, days_32, days_33, days_35), list(mean = mean, max = max))
  )

thresholds_ytd
```

```{r july-ytd-ranks}
july_ytd %>%
  arrange(desc(mean_tmax)) %>%
  select(year, mean_tmax, mean_tmin, days_30, days_31, days_32, days_33, days_35, max_tmax, rain_total) %>%
  slice(1:8)
```

The current July is **hot, but not the hottest July 1-15 in the archive**. It ranks fourth on mean daily maximum temperature over the first 15 days. What is more extreme is the *single-day* heat: the hottest day so far is the hottest July 1-15 day in the archive, and there is already one day at or above 33°C.

```{r july-ytd-summary}
current_ytd %>%
  transmute(
    year,
    mean_tmax = round(mean_tmax, 2),
    mean_tmin = round(mean_tmin, 2),
    max_tmax = round(max_tmax, 2),
    days_30, days_31, days_32, days_33, days_35,
    rain_total = round(rain_total, 2),
    rainy_days,
    rank_mean_tmax,
    rank_max_tmax,
    rank_days_33
  )
```

## Chart 1: July 1-15 heat through time

```{r chart-ytd, fig.width=12, fig.height=6.5}
ytd_plot_data <- july_ytd %>%
  mutate(
    label = if_else(year == current_year, paste0(year, "  ", round(mean_tmax, 2), "°C"), NA_character_)
  )

record_ytd <- july_ytd %>% arrange(desc(mean_tmax)) %>% slice(1)

p_ytd <- ggplot(ytd_plot_data, aes(year, mean_tmax)) +
  geom_line(color = "#7a7a7a", linewidth = 0.7, alpha = 0.65) +
  geom_point(color = "#7a7a7a", size = 1.7, alpha = 0.75) +
  geom_point(data = current_ytd, color = "#CD3333", size = 3.2) +
  geom_point(data = record_ytd, color = "#005566", size = 2.9) +
  geom_text(
    data = current_ytd,
    aes(label = paste0(round(mean_tmax, 2), "°C\nrank ", fmt_rank(rank_mean_tmax, nrow(july_ytd)))),
    hjust = -0.02,
    nudge_x = 0.35,
    vjust = 0.25,
    size = 3.3,
    fontface = "bold",
    color = "#CD3333",
    lineheight = 0.95
  ) +
  geom_text(
    data = record_ytd,
    aes(label = paste0(year, " record\n", round(mean_tmax, 2), "°C")),
    hjust = 1.02,
    nudge_x = -0.35,
    vjust = -0.15,
    size = 3.1,
    fontface = "bold",
    color = "#005566",
    lineheight = 0.95
  ) +
  scale_x_continuous(breaks = seq(1985, 2025, 5), limits = c(1981, 2028)) +
  scale_y_continuous(labels = label_number(accuracy = 0.1, suffix = "°C")) +
  labs(
    title = "July 1-15 heat is high, but not a record on average",
    subtitle = "Mean daily maximum temperature for 1-15 July in each year; 2026 is highlighted in red and the archive record in blue.",
    caption = "Hourly ERA5 archive for Bangalore"
  ) +
  theme_weather_question()

p_ytd

chart_ytd_path <- file.path(chart_dir, "july_hot_rainfall_ytd_heat.png")
ggsave(chart_ytd_path, p_ytd, width = 12, height = 6.5, dpi = 180, bg = "white")
```

## What does a hot July mean for Jul-Sep rainfall?

This is the historical full-July test. I use complete Julys from 1981-2025, then compare them with the total rainfall in July-September of the same year.

```{r july-full-rain-prep}
july_full <- daily_weather %>%
  filter(month == 7, year <= current_year - 1) %>%
  summarise(
    mean_tmax = mean(tmax),
    mean_tmin = mean(tmin),
    mean_tmean = mean(tmean),
    days_30 = sum(tmax >= 30),
    days_31 = sum(tmax >= 31),
    days_32 = sum(tmax >= 32),
    days_33 = sum(tmax >= 33),
    days_35 = sum(tmax >= 35),
    max_tmax = max(tmax),
    .by = year
  )

jas_rain <- daily_weather %>%
  filter(month %in% 7:9, year <= current_year - 1) %>%
  summarise(jas_rain_mm = sum(rain_mm), .by = year)

july_full <- july_full %>%
  left_join(jas_rain, by = "year") %>%
  mutate(heat_quartile = ntile(mean_tmax, 4))

cor_heat_rain <- cor(july_full$mean_tmax, july_full$jas_rain_mm, use = "complete.obs")
spearman_heat_rain <- cor(july_full$mean_tmax, july_full$jas_rain_mm, use = "complete.obs", method = "spearman")

list(
  pearson = cor_heat_rain,
  spearman = spearman_heat_rain
)
```

```{r heat-rain-quartiles}
july_full %>%
  group_by(heat_quartile) %>%
  summarise(
    n = n(),
    mean_heat = mean(mean_tmax),
    mean_jas_rain = mean(jas_rain_mm),
    median_jas_rain = median(jas_rain_mm),
    min_jas_rain = min(jas_rain_mm),
    max_jas_rain = max(jas_rain_mm),
    .groups = "drop"
  ) %>%
  mutate(
    mean_heat = round(mean_heat, 2),
    mean_jas_rain = round(mean_jas_rain, 1),
    median_jas_rain = round(median_jas_rain, 1),
    min_jas_rain = round(min_jas_rain, 1),
    max_jas_rain = round(max_jas_rain, 1)
  )
```

```{r chart-rain, fig.width=12, fig.height=6.5}
hot_full <- july_full %>% arrange(desc(mean_tmax)) %>% slice(1:4)

p_rain <- ggplot(july_full, aes(mean_tmax, jas_rain_mm)) +
  geom_point(color = "#7a7a7a", size = 2, alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "#005566", linewidth = 0.9, linetype = "dashed") +
  geom_text(
    data = hot_full,
    aes(label = year),
    nudge_y = 10,
    size = 3.2,
    fontface = "bold",
    color = "#3C3C3C"
  ) +
  annotate(
    "text",
    x = min(july_full$mean_tmax) + 0.05,
    y = max(july_full$jas_rain_mm),
    hjust = 0,
    vjust = 1,
    label = paste0(
      "Pearson r = ", round(cor_heat_rain, 2),
      "\nSpearman ρ = ", round(spearman_heat_rain, 2)
    ),
    size = 3.5,
    fontface = "bold",
    color = "#005566"
  ) +
  scale_y_continuous(labels = label_number(suffix = " mm")) +
  scale_x_continuous(labels = label_number(accuracy = 0.1, suffix = "°C")) +
  labs(
    title = "Hot Julys do not reliably predict wetter or drier Jul-Sep seasons",
    subtitle = "Complete Julys from 1981-2025 vs same-year July-September rainfall totals. The relationship is close to flat.",
    caption = "Hourly ERA5 archive for Bangalore"
  ) +
  theme_weather_question()

p_rain

chart_rain_path <- file.path(chart_dir, "july_hot_rainfall_seasonal_rain.png")
ggsave(chart_rain_path, p_rain, width = 12, height = 6.5, dpi = 180, bg = "white")
```

## Checks

```{r checks}
list(
  july_1_15_2026_rank_mean_tmax = current_ytd$rank_mean_tmax,
  july_1_15_2026_rank_max_tmax = current_ytd$rank_max_tmax,
  july_1_15_2026_rank_days_33 = current_ytd$rank_days_33,
  july_1_15_2026_mean_tmax = current_ytd$mean_tmax,
  july_full_heat_rain_pearson = cor_heat_rain,
  july_full_heat_rain_spearman = spearman_heat_rain
)
```

## Verdict

Insightful: no

The July 1-15 heat is genuinely elevated, but it is not a record on the month-to-date mean maximum temperature scale, and the hotter-end signal is mostly in the single-day peak rather than sustained extreme heat. More importantly, the historical full-July comparison does **not** show a meaningful relationship between hot Julys and Jul-Sep rainfall: the correlation is essentially flat, and the hottest quartile of Julys has about the same seasonal rainfall as the coolest quartile. That makes this a good descriptive check, but not strong enough to publish.

Charts:

- `docs/assets/analysis/july_hot_rainfall_ytd_heat.png`
- `docs/assets/analysis/july_hot_rainfall_seasonal_rain.png`

Blog post:

- not published
- if this ever becomes publishable, create `docs/blog/july-hot-rainfall.html`
- include `This post is AI-written.` at the top of the article body
- end with an `Original question` section containing `r rmarkdown::metadata$trigger_prompt`
- add the post to the top of `docs/blog/index.html` and `docs/blog/feed.xml`
- add `blog_url: "blog/july-hot-rainfall.html"` to the top of `analysis_questions/analyses.yml`
