---
title: "How bad was July 2026 in Bangalore?"
question: "How bad was this July in bangalore in terms of rainfall and heat ?"
trigger_prompt: |
  Work in /home/karthik/apps/bangalore-weather. HERMES DRAFT MODE. Treat the following as the exact weather question trigger_prompt. Follow the Hermes draft workflow in analysis_questions/README.md and AGENTS.md: create and run the analysis notebook, write the Verdict, create analysis charts if needed, and decide insightfulness strictly. Do not create or modify docs/blog, docs/index.html, docs/blog/feed.xml, analysis_questions/analyses.yml, do not commit, and do not push. Reply with the gist of the analysis, the insightfulness verdict, files created, and if publishable ask for explicit approval using /weather_publish <slug>. Question: How bad was this July in bangalore in terms of rainfall and heat ?
slug: "july-2026-heat-rain"
date: "`r Sys.Date()`"
analyst: "Codex"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 6,
  dpi = 160
)

library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)

root_dir <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
chart_dir <- file.path(root_dir, "docs", "assets", "analysis")
dir.create(chart_dir, recursive = TRUE, showWarnings = FALSE)

load(file.path(root_dir, "data", "bangaloreTemperature.RData"))
load(file.path(root_dir, "data", "bangaloreRainfall.RData"))

analysis_date <- max(as_datetime(blrTemp$DT), na.rm = TRUE) |> as_date()
analysis_year <- year(analysis_date)

n_years <- length(unique(year(as_datetime(blrTemp$DT))))

analysis_theme <- function(base_size = 11) {
  theme_minimal(base_size = base_size, base_family = "Helvetica") +
    theme(
      plot.background = element_rect(fill = "#e5e1d8", color = NA),
      panel.background = element_rect(fill = "#e5e1d8", color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
      axis.title = element_blank(),
      axis.text = element_text(color = "#3C3C3C", face = "bold"),
      plot.title = element_text(color = "#3C3C3C", face = "bold"),
      plot.subtitle = element_text(color = "#666666"),
      plot.caption = element_text(color = "#666666", size = rel(0.85)),
      plot.margin = margin(15, 18, 15, 15)
    )
}
```

## Question

`r rmarkdown::metadata$question`

## Analysis frame

- July 2026 is complete in the local archive, so the clean comparison is full July 2026 vs full historical Julys.
- "Heat" is measured by the monthly mean of daily maximum temperature.
- "Rainfall" is measured by total July precipitation.
- I also check how many July days crossed 30°C, 31°C, 32°C, 33°C, and 35°C because average heat can hide a lack of true spikes.
- Data source is the local ERA5 archive only.

## Data prep

```{r data-prep}
temp_hourly <- blrTemp %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    temp_c = as.numeric(Temp)
  ) %>%
  filter(!is.na(temp_c))

rain_hourly <- blrRain %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    rain_mm = as.numeric(Rain)
  ) %>%
  filter(!is.na(rain_mm))

daily_weather <- temp_hourly %>%
  summarise(
    tmin = min(temp_c, na.rm = TRUE),
    tmax = max(temp_c, na.rm = TRUE),
    tmean = mean(temp_c, na.rm = TRUE),
    .by = c(date, year, month)
  ) %>%
  left_join(
    rain_hourly %>%
      summarise(rain_mm = sum(rain_mm, na.rm = TRUE), .by = c(date, year, month)),
    by = c("date", "year", "month")
  ) %>%
  filter(is.finite(tmin), is.finite(tmax), !is.na(rain_mm))

july_stats <- daily_weather %>%
  filter(month == 7) %>%
  summarise(
    mean_tmax = mean(tmax),
    mean_tmin = mean(tmin),
    mean_tmean = mean(tmean),
    max_tmax = max(tmax),
    rain_total = sum(rain_mm),
    rainy_days = sum(rain_mm > 0.1),
    days_30 = sum(tmax >= 30),
    days_31 = sum(tmax >= 31),
    days_32 = sum(tmax >= 32),
    days_33 = sum(tmax >= 33),
    days_35 = sum(tmax >= 35),
    .by = year
  ) %>%
  mutate(
    rank_heat = min_rank(desc(mean_tmax)),
    rank_rain = min_rank(rain_total),
    rank_dry = min_rank(rain_total),
    heat_percentile = percent_rank(mean_tmax),
    rain_percentile = percent_rank(rain_total)
  )

current_july <- july_stats %>% filter(year == analysis_year)
current_july
```

## Quick facts

```{r facts}
july_stats %>%
  summarise(
    year = first(year),
    n_years = n(),
    current_mean_tmax = round(current_july$mean_tmax, 2),
    current_rain_total = round(current_july$rain_total, 1),
    heat_rank = paste0(current_july$rank_heat, " / ", n()),
    rain_rank = paste0(current_july$rank_rain, " / ", n()),
    heat_percentile = round(current_july$heat_percentile, 3),
    rain_percentile = round(current_july$rain_percentile, 3),
    max_tmax = round(current_july$max_tmax, 2),
    days_30 = current_july$days_30,
    days_31 = current_july$days_31,
    days_32 = current_july$days_32,
    days_33 = current_july$days_33,
    days_35 = current_july$days_35,
    rainy_days = current_july$rainy_days
  )
```

## Heat and rainfall through time

```{r chart-main, fig.width=12, fig.height=8.5}
heat_median <- median(july_stats$mean_tmax)
rain_median <- median(july_stats$rain_total)

heat_panel <- ggplot(july_stats, aes(year, mean_tmax)) +
  geom_line(color = "grey70", linewidth = 0.8) +
  geom_point(color = "grey65", size = 1.7) +
  geom_hline(yintercept = heat_median, linetype = "dashed", color = "#888888", linewidth = 0.4) +
  geom_point(data = current_july, color = "#CD3333", size = 3.2) +
  geom_text(
    data = current_july,
    aes(label = paste0(round(mean_tmax, 2), "°C\nrank ", rank_heat, " hottest / ", n_years)),
    hjust = -0.05,
    vjust = 0.5,
    nudge_x = 0.35,
    size = 3.3,
    fontface = "bold",
    color = "#CD3333",
    lineheight = 0.95
  ) +
  scale_x_continuous(breaks = seq(1985, 2025, 5), limits = c(1981, 2028)) +
  scale_y_continuous(labels = label_number(accuracy = 0.1, suffix = "°C")) +
  labs(
    title = "Mean July heat",
    subtitle = "Monthly mean of daily maximum temperature; 2026 is highlighted in red.",
    caption = "Hourly ERA5 archive for Bangalore"
  ) +
  analysis_theme()

rain_panel <- ggplot(july_stats, aes(year, rain_total)) +
  geom_line(color = "grey70", linewidth = 0.8) +
  geom_point(color = "grey65", size = 1.7) +
  geom_hline(yintercept = rain_median, linetype = "dashed", color = "#888888", linewidth = 0.4) +
  geom_point(data = current_july, color = "#CD3333", size = 3.2) +
  geom_text(
    data = current_july,
    aes(label = paste0(round(rain_total, 1), " mm\nrank ", rank_dry, " driest / ", n_years)),
    hjust = -0.05,
    vjust = 0.5,
    nudge_x = 0.35,
    size = 3.3,
    fontface = "bold",
    color = "#CD3333",
    lineheight = 0.95
  ) +
  scale_x_continuous(breaks = seq(1985, 2025, 5), limits = c(1981, 2028)) +
  scale_y_continuous(labels = label_number(suffix = " mm")) +
  labs(
    title = "Total July rainfall",
    subtitle = "Monthly precipitation total; lower is drier and 2026 is highlighted in red.",
    caption = "Hourly ERA5 archive for Bangalore"
  ) +
  analysis_theme()

combined <- heat_panel / rain_panel +
  plot_annotation(
    title = "July 2026 was hot *and* dry by Bangalore standards",
    subtitle = paste0(
      "Historical July comparison from 1981-2026. 2026 ranks ",
      current_july$rank_heat, "th hottest on mean daily maximum temperature and ",
      current_july$rank_rain, "th driest on rainfall total."
    ),
    caption = "Hourly ERA5 archive for Bangalore"
  )

combined

chart_path <- file.path(chart_dir, "july_2026_heat_rain.png")
ggsave(chart_path, combined, width = 12, height = 8.5, dpi = 180, bg = "#e5e1d8")
```

## Checks

```{r checks}
list(
  analysis_date = analysis_date,
  july_2026_heat_rank = current_july$rank_heat,
  july_2026_rain_rank = current_july$rank_rain,
  july_2026_mean_tmax = current_july$mean_tmax,
  july_2026_rain_total = current_july$rain_total,
  july_2026_days_30 = current_july$days_30,
  july_2026_days_31 = current_july$days_31,
  july_2026_days_32 = current_july$days_32,
  july_2026_days_33 = current_july$days_33,
  july_2026_days_35 = current_july$days_35,
  july_heat_rain_correlation = cor(july_stats$mean_tmax, july_stats$rain_total, use = "complete.obs"),
  july_heat_rain_spearman = cor(july_stats$mean_tmax, july_stats$rain_total, use = "complete.obs", method = "spearman")
)
```

## Verdict

Insightful: no

This July was genuinely rough: it was the **4th hottest** July in the archive on mean daily maximum temperature and the **5th driest** on rainfall. But the heat was more persistent than extreme — the hottest day only reached 30.68°C, and there were no 31°C+ or 33°C+ days — so the month reads as a warm, dry monsoon-month rather than a standout heatwave.

That makes it a good factual summary, but not strong enough to publish as a site post. The result is descriptive rather than surprising, and the chart mostly confirms the ranking story in one glance.

Charts:

- `docs/assets/analysis/july_2026_heat_rain.png`

Blog post:

- not published
- if this ever becomes publishable, create `docs/blog/july-2026-heat-rain.html`
- include `This post is AI-written.` at the top of the article body
- end with an `Original question` section containing `r rmarkdown::metadata$trigger_prompt`
- add the post to the top of `docs/blog/index.html` and `docs/blog/feed.xml`
- add `blog_url: "blog/july-2026-heat-rain.html"` to the top of `analysis_questions/analyses.yml`
