---
title: "Question analysis template"
question: "Replace with the weather question"
slug: "replace-with-short-slug"
date: "`r Sys.Date()`"
analyst: "Codex"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 6,
  dpi = 150
)

library(tidyverse)
library(lubridate)
library(scales)

root_dir <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
chart_dir <- file.path(root_dir, "docs", "assets", "analysis")
dir.create(chart_dir, recursive = TRUE, showWarnings = FALSE)

load(file.path(root_dir, "data", "bangaloreTemperature.RData"))
load(file.path(root_dir, "data", "bangaloreRainfall.RData"))

if (file.exists(file.path(root_dir, "data", "bangaloreWind.RData"))) {
  load(file.path(root_dir, "data", "bangaloreWind.RData"))
}

theme_weather_question <- function(base_size = 11) {
  theme_minimal(base_size = base_size, base_family = "Helvetica") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
      axis.title = element_blank(),
      axis.text = element_text(color = "#3C3C3C", face = "bold"),
      plot.title = element_text(color = "#3C3C3C", face = "bold"),
      plot.subtitle = element_text(color = "#666666"),
      plot.caption = element_text(color = "#666666", size = rel(0.85)),
      legend.position = "none"
    )
}
```

## Question

`r rmarkdown::metadata$question`

## Data Prep

```{r data-prep}
temp_hourly <- blrTemp %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    doy = yday(DT),
    hour = hour(DT),
    temp_c = as.numeric(Temp)
  ) %>%
  filter(!is.na(temp_c))

rain_hourly <- blrRain %>%
  transmute(
    DT = as_datetime(DT),
    date = as_date(DT),
    year = year(DT),
    month = month(DT),
    doy = yday(DT),
    hour = hour(DT),
    rain_mm = as.numeric(Rain)
  ) %>%
  filter(!is.na(rain_mm))

daily_weather <- temp_hourly %>%
  summarise(
    tmin = min(temp_c, na.rm = TRUE),
    tmax = max(temp_c, na.rm = TRUE),
    tmean = mean(temp_c, na.rm = TRUE),
    .by = c(date, year, month, doy)
  ) %>%
  left_join(
    rain_hourly %>%
      summarise(rain_mm = sum(rain_mm, na.rm = TRUE), .by = c(date, year, month, doy)),
    by = c("date", "year", "month", "doy")
  ) %>%
  filter(is.finite(tmin), is.finite(tmax), !is.na(rain_mm))

daily_weather %>%
  summarise(
    first_date = min(date),
    last_date = max(date),
    days = n(),
    years = n_distinct(year)
  )
```

## Analysis

Replace this section with the smallest analysis that answers the question. Prefer
transparent summaries over clever models.

```{r starter-summary}
daily_weather %>%
  summarise(
    mean_tmax = mean(tmax),
    mean_tmin = mean(tmin),
    mean_annual_rain = sum(rain_mm) / n_distinct(year)
  )
```

## Chart

Save any public chart to `docs/assets/analysis/` with a stable, descriptive filename.

```{r starter-chart}
chart_path <- file.path(chart_dir, paste0(rmarkdown::metadata$slug, ".png"))

p <- daily_weather %>%
  summarise(rain_mm = sum(rain_mm), .by = year) %>%
  ggplot(aes(year, rain_mm)) +
  geom_line(color = "#005566", linewidth = 0.8) +
  geom_point(color = "#005566", size = 1.6) +
  scale_y_continuous(labels = label_number(suffix = " mm")) +
  labs(
    title = "Replace this starter chart",
    subtitle = "Annual rainfall, used here only as a template placeholder",
    caption = "Hourly Oikolab/ERA5 archive for Bangalore"
  ) +
  theme_weather_question()

p

# Uncomment only when the chart is final enough to publish or review.
# ggsave(chart_path, p, width = 12, height = 6, dpi = 180, bg = "white")
```

## Checks

List sensitivity checks here: alternate thresholds, excluded years, seasonal boundaries,
or any reason the answer might be fragile.

## Verdict

Insightful: no

This template does not answer a real question yet. Do not add it to the site.

Charts:

- none
